---
title: "Protocol for a Systematic Review"
description: |
  Effects of landfill location on human health
author:
  - name: Alex Vand
    url: {}
date: 2021-09-09
output:
  distill::distill_article:
    self_contained: false
categories:
  - environmental justice
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Abstract

## Study background

With modern consumerism, the accumulation of municipal waste and how to safely manage it has become a pressing environmental concern. There are over 13,000 active and inactive landfills in the United States, some located near residential areas. These landfills have the potential to release toxic substances into the air and groundwater, negatively affecting the health and well-being of nearby populations. This raises an important research question which will be addressed in this systematic review: How does the location of a solid waste facility affect the health of humans that live nearby? While a similar systematic review was conducted in 2009 by Porta et. al., a significant amount of research has been published in the past decade, prompting a more recent review to capture the new studies as well. Furthermore, because waste facilities are located disproportionately in minority areas, this systematic review will address this question within the context of prioritizing environmental justice as we continue to build sustainable communities.

## Summary of Study Methods

Our team used the PICO/PECO framework in order to formulate our question, where the population of interest is humans; the intervention or exposure is living near a landfill; the comparison is other humans that do not live near a landfill; and the outcome is the effect on human health. In order to find published research regarding this question, we conducted a search on the Web of Science bibliographic database using Boolean operators and wildcard indicators. For the search string itself, various synonyms for landfills and waste facilities were used and several different measures of human health, both physical and mental, were accounted for. Ultimately, our search yielded 5,716 articles. Each title will be inspected in the systematic review, and abstracts and entire texts will be examined as needed in order to determine eligibility for inclusion. The summary statistics from each publication will be tabulated and ultimately presented in a forest plot.

# Keywords

Landfill, waste facility, human health

# Background

As global consumerism continues to increase, the widespread use of single-use plastics and fast fashion contribute significant amounts of waste in today’s society. According to the Environmental Protection Agency (EPA), about 300 million tons of trash were generated in the United States in 2018, corresponding to nearly 5 pounds per person per day. This waste must be stored somewhere, and there are over 3,000 active landfills and 10,000 inactive landfills in the United States, some of which are located in close proximity to residential areas.

Title 40 of the Code of Federal Regulations (CFR) part 258 lists the restrictions for a municipal solid waste landfill (MSWLF) in the United States, including location criteria, composite liner requirements, leachate collection and removal systems, groundwater monitoring requirements, among others. While the United States has created some effective measures to sustain safe landfilling, many developing nations do not have the same infrastructure in place. As shown in Figure 1 below, the repercussions of residing near a waste facility may have negative impacts on human health. Unfortunately, the “management of solid waste (mainly landfills and incineration) releases a number of toxic substances” (Njoku et al. 2019). One notable substance is ambient toxic hydrogen sulfide (H2S), which causes malodor and “has been shown to negatively impact the health and well-being of populations” (Heaney et al. 2011). Additionally, “the by-products of solid waste deposited in a landfill have adverse effects on the surrounding environment and humans living closer to landfill sites” (Njoku et. al. 2019), leading to increased risk of congenital anomalies and low birth weight for populations living within two kilometers of landfills (Porta et. al. 2009).

Notably, solid waste facilities are located disproportionately in minority areas (Anderton et al. 1994), specifically in communities of color and low socioeconomic status (Heaney et al. 2011). Addressing the question of how landfills affect human health is critical to environmental science as we decide how to implement future waste facilities across the globe. Perhaps most importantly, we must prioritize environmental justice as we continue to build sustainable communities and minimize the disproportionate impacts on underserved communities around the world.


```{r}
library(DiagrammeR)

grViz("
      
digraph {

layout = dot

# add node statement

node [shape = circle]
A[label = 'Residing near a MSWLF'];
B[label = 'Negative human health impacts
on mental and physical well-being']

# add node statement
node [shape = box]
1[label = 'Increased exposure to toxi substances'];
2[label = 'Higher levels of H2S leading to malodor'];
3[label = 'Groundwater contamination']


# add edge statement
A->1; A->2; A->3;

1->B; 2->B; 3->B;

}      
")
```



Figure 1. Conceptual model indicating how living near a municipal solid waste landfill (MSWLF) might lead to negative impacts on human health.

# Objective of the review

## Primary research question
The primary research question that this systematic review protocol focuses on is: **what is the effect of landfill location on human health?** This question explores the effects of solid waste facilities on human health, including both mental and physical wellbeing. This systematic review does not examine the effects of the disposal site on flora, fauna, or the surrounding ecosystem.

## Components of primary research question (PICO)

The primary research question components are:

- **Population**: humans

- **Intervention**/Exposure: living near a landfill or solid waste facility

- **Comparator**: humans that do not live near a waste facility

- **Outcome**: effect on human health


# Methods

## Search string

The complete search string including Boolean operators and wildcards is:

(human$ OR people$ OR population$ OR communit* OR town* OR cit*) AND (landfill* OR “waste facilit*” OR “waste dump” OR “dumping ground” OR dump* OR “disposal area”) AND (health OR well$being OR wellness OR happ* OR “mental health” OR “physical health” OR safe*)


### TABLE


## Bibliographic databases

For this systematic review, Web of Science was used as the bibliographic database, including the institution subscriptions listed below in Table 2. 


### TABLE

## Search language(s)

The primary language used to complete this search is English. While this search was not explicitly restrictive to the English language, all search terms are in English; therefore the vast majority of the search results are also likely to be in English. Due to our limited selection of search languages, there is inherent bias in the search and search results as it excludes studies conducted in other languages and likely in other cultural contexts. Articles written in Arabic, Chinese, French, Russian, and Spanish may be relevant to this research question as these languages (along with English) are the six working languages among the United Nations community. However, it may also be relevant to include studies conducted in additional languages in order to account for case studies in developing countries where solid waste facilities have a significant impact on public health. A successful expansion of this search would require collaboration with multilingual scholars who have domain expertise as well as data science knowledge.


## Websites of relevant organizations and supplementary internet searches


### TABLE 

## Benchmark studies to estimate search comprehensiveness

This following benchmark publication is included in the Web of Science bibliographic search.

Gouveia, Nelson, and Rogerio Ruscitto do Prado. “Health Risks in Areas Close to Urban Solid Waste Landfill Sites.” *Revista de Saúde Pública* 44 (October 2010): 859–66. https://doi.org/10.1590/S0034-89102010005000029.

# Article screening and study eligibility criteria
 
## Screening process
In this review, we will examine the title of all search results. If the title is insufficient, the abstract will be examined in order to determine eligibility. We will use SysRev to compile results from our literature search. Articles will initially be categorized as (a) include, (b) exclude, and (c) maybe include. Our team will read the articles in full within category (c) in order to determine whether they are appropriate to include. We will evaluate inter-reviewer consistency via the kappa test, with some overlap in article review in order to determine any discrepancies. SysRev will be instrumental in recording whether or not articles meet the inclusion criteria. Inconsistent categorization among different reviewers of the same article will be resolved through a discussion and consensus among the reviewers of that article.

## Eligibility criteria applied in screening process

### Population (P) criteria
The populations studied must be human subjects and may not include any nonhuman species. Studies focused on plants, animals, ecosystems, and agriculture are not eligible for inclusion.

### Intervention (I) criteria
The intervention or exposure of interest is whether the population is located near a landfill or solid waste facility. Studies focused on the effects of radioactive or hazardous waste facilities are not considered for this systematic review.

### Comparator (C) criteria
The comparator for this research question is other human populations that do not live near a solid waste facility.

### Outcome (O) criteria
The outcome component explores the effects on human mental and physical health and wellbeing as related to exposure to solid waste facilities. This systematic review excludes studies focused on congenital conditions that are unrelated to landfill location.

### Study design criteria
Study designs include spatial comparisons of two groups (subjectively distal and proximal to the waste facility), case studies for specific locations, and available health records.

### Language criteria
Although no languages are explicitly excluded, English is the only language included as we are lacking in multilingual collaborators.

### Tabulated screening criteria

### TABLE 4

## Documentation of eligibility
We will include a full list of all reviewed articles that have been excluded from the systematic review. The citations of the excluded articles will be documented along with the rationale for exclusion.

#     Data coding and extraction strategy
## Data extraction

In an ideal case where the paper states the effect size and uncertainty of the study, we will tabulate these statistical values and include this information in the systematic review. For non-ideal cases where the raw data exists but the summary statistics have not yet been calculated, our team will extract the underlying primary data and calculate the average values and standard deviations ourselves.

## Potential explanatory variables for heterogeneity among studies

In order to account for heterogeneity among studies, our team will compile potential explanatory variables along with each of the publications. Potential explanatory variables include population variables that account for social demographics, such as socioeconomic status and race. Income, property value, and educational attainment could be considered indications of wealth and socioeconomic status. While race and educational attainment are categorical variables, household income and property value are continuous and quantitative (unless these values are “binned” into specific intervalled categories). Demographic variables, such as age, race, income, gender, and lifestyle play a major role in determining human health and wellbeing and this review would be lacking if we did not explore the statistical relationships among these variables and the data gathered from our search.

## Data presentation

A forest plot (or blobbogram) will be constructed to display the findings of the meta-analysis in order to capture the results from many studies and summarize these findings into one schematic. Additionally, a map of landfill locations overlaid with a map of health statistics may be an effective presentation of the results of this meta-analysis.

# Competing interests
The author declares no competing interests with the proposed research project.

# References
Christopher D. Heaney et al., “Relation between Malodor, Ambient Hydrogen Sulfide, and Health in a Community Bordering a Landfill,” *Environmental Research* 111, no. 6 (August 1, 2011): 847–52, https://doi.org/10.1016/j.envres.2011.05.021.

Daniela Porta et al., “Systematic Review of Epidemiological Studies on Health Effects Associated with Management of Solid Waste,” *Environmental Health* 8, no. 1 (December 23, 2009): 60, https://doi.org/10.1186/1476-069X-8-60.

Douglas L. Anderton et al., “Environmental Equity: The Demographics of Dumping,” Demography 31, no. 2 (May 1, 1994): 229–48, https://doi.org/10.2307/2061884.

Nelson Gouveia and Rogerio Ruscitto do Prado, “Health Risks in Areas Close to Urban Solid Waste Landfill Sites,” *Revista de Saúde Pública* 44 (October 2010): 859–66, https://doi.org/10.1590/S0034-89102010005000029.

Prince O. Njoku, Joshua N. Edokpayi, and John O. Odiyo, “Health and Environmental Risks of Residents Living Close to a Landfill: A Case Study of Thohoyandou Landfill, Limpopo Province, South Africa,” *International Journal of Environmental Research and Public Health* 16, no. 12 (January 2019): 2125, https://doi.org/10.3390/ijerph16122125.





